Static contract checking for Haskell

### 2011-10-20 paper.tex
Paper: partial update of section 3.

figure~\ref{dtrans-reg} gives the translation of function not defined by
pattern matching, which is really easy: we just have to state the
equality between le left-hand side and the right-hand side.

Translating definitions that use pattern-matching is more challenging
and is described in figure~\ref{dtrans}.

The first line says that when applied to an argument that matches a
pattern of the case expression, we should equate the function call to
the corresponding expression.

The second line states that if the pattern-matching failed or if we
pattern-matched on $\bad$ then the result should be $\unr$.

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1, \dots, K_n} &=& \ktrans{\mbox{data } T = K_1, \dots, K_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = \etrans{e} \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& a = \etrans{e_1} \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n} \\%removed } here ...
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

\subsection{$\ktrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
  on its $k$-th component.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ None of the $K_i$ is unreachable.
\item One may want to also state that if $\vec{x} \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\cf{\vec{x}} \to \cf{K_i(\vec{x})}$.
\end{itemize}

>>>>

figure~\ref{dtrans} gives the translation of function not defined by
pattern matching, which is really easy: we just have to state the
equality between le left-hand side and the right-hand side.

Translating definitions that use pattern-matching is more challenging
and is described in figure~\ref{dtrans}.  The first line says that
when applied to an argument that matches a pattern of the case
expression, we should equate the function call to the corresponding
expression. The second line states that if the pattern-matching failed
or if we pattern-matched on $\bad$ then the result should be
$\unr$.\note[nc]{This text does not agree with the figure.  In
  particular, the axiom says nothing about the result of scrutinizing
  $\bad$.  I think the correct behavior is that scrutinizing $\bad$
  results in $\bad$, \emph{not} $\unr$.}

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \Phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} &=& \ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f(\vec{x}) = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  f(\vec{x})) \to a = \etrans{e} \\
   \raisebox{ 0.5 cm }{$\dtrans{f(\vec{x}) = \mbox{case } e \mbox{ of } [K_i(\vec{x_i}) = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = f(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = K_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~\etrans{e} = K_1(\vec{x_1}) &\to& a = \etrans{e_1} \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~\etrans{e} = K_n(\vec{x_n}) &\to& a = \etrans{e_n} \\%removed } here ...
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

\subsection{$\ktrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in figure~\ref{ktrans}
$$\ktrans{\mbox{data } T = K_1 a_1, \dots, K_n a_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
    ``type'' is used intuitively here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ Now full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
\end{itemize}
